Perceptions of Inequality and Meritocracy: Their Interplay in Shaping Preferences for Market Justice in Chile (2016-2023)

### 2025-04-23 03-methods.qmd
Update paper

---
author: "Equipo EDUMER"
bibliography: "input/bib/refs.bib"
csl: "input/bib/apa6.csl"
---

@@

@@

@@

>>>>

---
author: "Equipo EDUMER"
bibliography: "input/bib/merit-edjust.bib"
csl: "input/bib/apa6.csl"
---

```{r}
#| echo: false
#| message: false
#| warning: false

library(knitr)
knitr::opts_chunk$set(echo = TRUE, include = TRUE, warning = FALSE, message = FALSE)

table_format <- if(is_html_output()) {
  "html"
} else if(is_latex_output()) {
  "latex"
}
table_format2 <- if(is_html_output()) {
  T
} else if(is_latex_output()) {
  F
}

options(kableExtra.html.bsTable = T)
options(knitr.kable.NA = "")
```

```{r}

#| include: false
#| echo: false

if (! require("pacman")) install.packages("pacman")

pacman::p_load(tidyverse, 
               sjmisc, 
               sjPlot, 
               lme4, 
               here, 
               performance,
               influence.ME, 
               marginaleffects,
               MLMusingR,
               texreg, 
               ggdist,
               misty,
               kableExtra,
               ggalluvial, 
               shadowtext,
               MetBrewer,
               patchwork,
               sjlabelled,
               summarytools)

options(scipen=999)
rm(list = ls())
```

```{r}
#| echo: false
#| include: false

load(file = here("input/data/proc/df_study1_long_t7.RData"))

glimpse(df_study1_long_t7)

# Generate analytical sample

df_study1 <- df_study1_long_t7 %>%
  select(idencuesta,
         ola,
         ponderador_long_total,
         segmento,
         estrato,
         just_health,
         just_pension,
         just_educ,
         mjp, 
         perc_inequality, 
         just_inequality,
         merit_effort, 
         merit_talent, 
         educ,
         quintil1,
         sex,
         age,
         ess, 
         ideo) %>% 
  na.omit() %>% 
  mutate(ola = as.factor(ola),
         ola_num = as.numeric(ola),
         ola_2=as.numeric(ola)^2)

df_study1 <- df_study1 %>%
  group_by(idencuesta) %>%             # Agrupar por el identificador del participante
  mutate(n_participaciones = n()) %>%  # Contar el número de filas (participaciones) por participante
  ungroup()

df_study1 <- df_study1 %>% filter(n_participaciones>1)

# Corregir etiquetas

df_study1$just_health <- sjlabelled::set_label(df_study1$just_health, 
                        label = "Health distributive justice")

df_study1$just_pension <- sjlabelled::set_label(df_study1$just_pension, 
                        label = "Pension distributive justice")

df_study1$just_educ <- sjlabelled::set_label(df_study1$just_educ, 
                        label = "Education distributive justice")

df_study1$merit_effort <- sjlabelled::set_label(df_study1$merit_effort, 
                        label = "People are rewarded for their efforts")

df_study1$merit_talent <- sjlabelled::set_label(df_study1$merit_talent, 
                        label = "People are rewarded for their intelligence")

```

@@

This study draws on data from the Chilean Longitudinal Social Survey (ELSOC), a panel study collected annually from 2016 to 2023. The survey evaluates how individuals think, feel, and behave regarding conflict and social cohesion in Chile. ELSOC employs a probabilistic, stratified, clustered, multistage sampling design encompassing major urban centers (Santiago, Valparaíso, and Concepción) and smaller cities. The target population includes women and men aged 18--75 who are habitual residents of private dwellings. The first wave included 2,927 participants, representing northern and southern regions, covering 77% of the country's total population and 93% of its urban population, with a response rate of 62.4% [@elsoc_estudio_2022].

The survey has been conducted yearly since 2016, except in 2020, when it was suspended due to the pandemic. Waves 2016, 2017, 2018, 2019, 2022, and 2023 used computer-assisted personal interviewing (CAPI), while a reduced wave in 2021 employed computer-assisted telephone interviews (CATI). Wave 3 included a refreshment sample (N = 1,519) to counter attrition, but these cases are not used here to capture longer response trends. Between wave 1 and wave 7, attrition amounted to 40%, achieving a final sample of N = 1,741. Longitudinal weights adjust for both the sampling design and potential biases arising from systematic non-participation. Further details on sampling, attrition, and weighting can be found at <https://coes.cl/encuesta-panel/>, and the dataset is publicly available at <https://dataverse.harvard.edu/dataverse/elsoc>.

@@

*Dependent variable*

**Market justice preferences**: The dependent variable in this study is preferences for market justice. This construct is operationalized through three items that capture how strongly individuals justify conditioning access to social services---healthcare, pensions, and education---on income. Specifically, the justification of inequality in healthcare is assessed by the question: "Is it fair in Chile that people with higher incomes can access better healthcare than people with lower incomes?" The same question is posed for pensions and education. In all cases, respondents indicate their level of agreement on a five-point Likert scale ranging from 1 ("strongly disagree") to 5 ("strongly agree"). Additionally, we include a composite measure of "market justice preferences", calculated as the average of these three items ($\alpha$ = 0.84). This index ranges from 1 to 5, with higher values indicating stronger preferences for market justice (see @tbl-summary1).

```{r}
#| label: tbl-summary1
#| tbl-cap: "Dependent variables for the first wave (2016)"
#| tbl-cap-location: top
#| results: asis
#| echo: false

t1 <- df_study1 %>% 
  filter(ola == 1) %>% 
  select(just_health, just_pension, just_educ, mjp)

df<-dfSummary(t1,
               plain.ascii = FALSE,
               style = "multiline",
               tmp.img.dir = "/tmp",
               graph.magnif = 0.75,
               headings = F,  # encabezado
               varnumbers = F, # num variable
               labels.col = T, # etiquetas
               na.col = F,    # missing
               graph.col = F, # plot
               valid.col = T, # n valido
               col.widths = c(30,10,10,10))

df$Variable <- NULL # delete variable column

print(df)
```

*Independent variables*

**Perception of economic inequality**: The main independent variable refers to the perception of economic inequality, measured through the perceived wage gap, which serves as an absolute indicator of inequality perceptions [@castillo_perception_2022]. This measure is derived from the gap between the perceived salaries of jobs at opposite ends of the occupational hierarchy. Specifically, it relies on the logarithm of the division between the perceived salary of a large-company president and that of an unskilled worker [@Castillo2011]. Higher values of this term indicate a greater perception of economic inequality between occupations located at the extremes of the status continuum.

**Justification of economic inequality**: This construct is measured as the just earnings gap, an approach to measuring the justification of economic inequality based on earning differences considered just between high and low status occupations [@Castillo2011]. Similarly to economic inequality perception, this indicator relies on the logarithmic ratio between the just salary of a large-company president and that of an unskilled worker. Higher values of this term indicate a greater justification of economic inequality.

**Meritocracy**: Meritocratic perception is operationalized through two components: one addressing effort and another focusing on talent [@young_rise_1962]. The item used to gauge effort is: "In Chile, people are rewarded for their efforts," while the item for talent is: "In Chile, people are rewarded for their intelligence and skills." In both cases, respondents indicate their level of agreement on a five-point Likert scale, ranging from 1 ("strongly disagree") to 5 ("strongly agree").

Table 2 shows the independent variables used, their response categories and their frequencies.

```{r}
#| label: tbl-summary2
#| tbl-cap: "Independent variables for the first wave (2016)"
#| tbl-cap-location: top
#| results: asis
#| echo: false

t2 <- df_study1 %>% 
  filter(ola == 1) %>% 
  select(perc_inequality, just_inequality, merit_effort, merit_talent)

df<-dfSummary(t2,
               plain.ascii = FALSE,
               style = "multiline",
               tmp.img.dir = "/tmp",
               graph.magnif = 0.75,
               headings = F,  # encabezado
               varnumbers = F, # num variable
               labels.col = T, # etiquetas
               na.col = F,    # missing
               graph.col = F, # plot
               valid.col = T, # n valido
               col.widths = c(30,10,10,10))

df$Variable <- NULL # delete variable column

print(df)
```

*Controls*

Sociodemographic and attitudinal variables are included to control for potential composition effects in the population. In terms of sociodemographic characteristics, we incorporate per capita household income quintile, educational level (1=Less than Universitary, 2=Universitary), age (in years), and sex (1=Male, 2=Female), which have been shown to influence market justice preferences significantly [@lindh_public_2015; @castillo_socialization_2024]. Regarding attitudinal variables, we include political identification (1=Left, 2=Center, 3=Right, 4=No identification) and subjective social status (ranging from 1 to 10) because they may confound the relationship between market justice preferences and perceptions of inequality and meritocracy [@schneider_poverty_2015].

@@

Given the data's hierarchical structure, in which observations are nested in survey waves, we employ longitudinal multilevel linear models [@singer_applied_2009]. In a panel-data framework, within-person effects capture how shifts in individual-level variables across waves are associated with variations in market justice preferences. By contrast, between-person effects focus on differences among individuals, explaining how long-term (or average) values relate to overall levels of market justice preferences.

To estimate within-person effects, we use group-mean centering, where each respondent functions as the "group" (i.e., observations nested within persons). Meanwhile, the between-person effects are derived from each individual's average on these variables, calculated across the waves of panel data.

All the analyses were conducted using R software and the *lme4* package [@bates_fitting_2015].

### 2025-04-23 paper.qmd
Update paper

---
title: "Perceptions of Inequality and Meritocracy: Their Interplay in Shaping Preferences for Market Justice in Chile (2016-2023)"
format:
  docx:
    number-sections: true
  html: 
    code-fold: true
    html-math-method: katex
    number-sections: true
  pdf: 
    geometry: 
      - top=30mm
      - left=20mm
    template-partials: 
      - title.tex
    keep-tex: true
    include-in-header:
      text: |
        \usepackage[noblocks]{authblk}
        \renewcommand*{\Authsep}{, }
        \renewcommand*{\Authand}{, }
        \renewcommand*{\Authands}{, }
        \renewcommand\Affilfont{\small}
    number-sections: true
editor: source
bibliography: "input/bib/refs.bib"
csl: "input/bib/apa6.csl"
author:
  - name: Juan Carlos Castillo
    corresponding: true
    orcid: 0000-0003-1265-7854
    email: juancastillov@uchile.cl
    affiliations:
      - ref: 1
      - ref: 2
      - ref: 3
  - name: Andreas Laffert
    orcid: 0000-0002-9008-2454
    email: alaffertt@estudiante.uc.cl 
    affiliations:
      - ref: 4
  - name: Kevin Carrasco
    orcid: 0000-0002-9348-8361
    email: kevin.carrasco@ug.uchile.cl
    affiliations:
      - ref: 2
  - name: Julio Iturra
    orcid: 0000-0003-1993-3479
    email: julioiturrasanhueza@gmail.com 
    affiliations:
      - ref: 5     
affiliations: 
  - id: 1
    name: Departamento de Sociología, Universidad de Chile
  - id: 2
    name: Centro de estudios del conflicto y cohesión social (COES)
  - id: 3
    name: Núcleo milenio de desigualdades y oportunidades digitales (NUDOS)
  - id: 4
    name: Instituto de Sociología, Pontificia Universidad Católica de Chile
  - id: 5
    name: International Graduate School of Social Sciencies (BIGSSS), University of Bremen, Germany
citeproc: true
abstract: | 
  My abstract. \newline
  **Keywords**: meritocracy, social inequality, inequality justification, COVID-19
link-citations: true
linestretch: 1.15       
mainfont: Times New Roman
fontsize: 12pt          
colorlinks: true
fig-height: 4
fig-width: 7.5
header-includes:
  - \usepackage{threeparttable}
---

@@

{{< include 07-supplementary-material.qmd >}}

>>>>

---
title: "Perceptions of Inequality and Meritocracy: Their Interplay in Shaping Preferences for Market Justice in Chile (2016-2023)"
format:
  docx:
    number-sections: true
  html: 
    code-fold: true
    html-math-method: katex
    number-sections: true
  pdf: 
    geometry: margin=2cm
    template-partials: 
      - title.tex
    keep-tex: true
    include-in-header:
      text: |
        \usepackage[noblocks]{authblk}
        \renewcommand*{\Authsep}{, }
        \renewcommand*{\Authand}{, }
        \renewcommand*{\Authands}{, }
        \renewcommand\Affilfont{\small}
    number-sections: true
editor: source
bibliography: "input/bib/merit-edjust.bib"
csl: "input/bib/apa6.csl"
author:
  - name: Juan Carlos Castillo
    corresponding: true
    orcid: 0000-0003-1265-7854
    email: juancastillov@uchile.cl
    affiliations:
      - ref: 1
      - ref: 2
      - ref: 3
  - name: Andreas Laffert
    orcid: 0000-0002-9008-2454
    email: alaffertt@estudiante.uc.cl 
    affiliations:
      - ref: 4
  - name: Kevin Carrasco
    orcid: 0000-0002-9348-8361
    email: kevin.carrasco@ug.uchile.cl
    affiliations:
      - ref: 2
  - name: Julio Iturra
    orcid: 0000-0003-1993-3479
    email: julioiturrasanhueza@gmail.com 
    affiliations:
      - ref: 5     
affiliations: 
  - id: 1
    name: Departamento de Sociología, Universidad de Chile
  - id: 2
    name: Centro de estudios del conflicto y cohesión social (COES)
  - id: 3
    name: Núcleo milenio de desigualdades y oportunidades digitales (NUDOS)
  - id: 4
    name: Instituto de Sociología, Pontificia Universidad Católica de Chile
  - id: 5
    name: International Graduate School of Social Sciencies (BIGSSS), University of Bremen, Germany
citeproc: true
abstract: | 
  My abstract. \newline
  **Keywords**: meritocracy, social inequality, inequality justification, COVID-19
link-citations: true
linestretch: 1.15       
mainfont: Times New Roman
fontsize: 12pt          
colorlinks: true
fig-height: 4
fig-width: 7.5
header-includes:
  - \usepackage{threeparttable}
---

@@

# References

::: {#refs}
:::
